---
title: "Basic R demo"
author: "Wing-Ho Ko"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Arithmetics

Using R as a calculator (note the operator precedence):

```{r}
3 + 4 * 2
```

The power operator. Again note the operator precedence

```{r}
5 + 4 * 3 ^ 2
```

Divisions are generally true division

```{r}
1 / 3
```

To do floor division (a.k.a. integer division) and to obtain the remainder (a.k.a. modulus), use `%/%` and `%%`

```{r}
100 %/% 3
```

```{r}
100 %% 3
```

Define numerical values using engineering notation

```{r}
5e3 + 200
```

### String operation

Defining character strings using both `'` and `"`

```{r}
a <- "apple"
b <- 'banana'
```

Concatenation with inserted space:

```{r}
paste(a, b)
```

Concatenation without space inserted:

```{r}
paste0(a, b)
```

### Variable assignment

Assign variables to numerical values, and perform basic arithmetic

```{r}
a <- 5
b <- 3

a * b
```

Variables assigned to the same value are still independent of each others

```{r}
a <- 5
b <- 5 
c <- a

a <- 3
b <- -1

print(a)
print(b)
print(c)
```

### Using functions

Calling a function using positional arguments

```{r}
log(4, 2)
```

Calling a function with both positional and named arguments

```{r}
log(4, base=2)
```

Calling a function with named argument (note that order **does not matter** in this case)

```{r}
log(base=2, x=4)
```

Use `c()` to combine two numbers into a vector

```{r}
a <- 5
b <- 3

y <- c(a, b)
print(y)
```

Create a sequence from 1 to 3, with step of 0.5

```{r}
seq(1, 3, 0.5)
```

Create a linear sequence from 1 to 3, with a total length of 10

```{r}
seq(1, 3, length.out=10)
```

