---
title: "R Dataframe Demo"
author: "Wing-Ho Ko"
output:
  html_notebook: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Loading packages for use

We'll load `readr` for importing external data as dataframes, `dplyr` for dataframe manipulation, and `car` for random sampling of dataframe.

When a package is loaded using the `library()` function, the functions it defines become available in the workspace and can be called directly. 

```{r}
library(readr)
library(dplyr)
library(car)
```

### Loading external data into a dataframe

As a demonstration, we'll load the Palmer penguins dataset in both online and offline manner.

To load the file directly from the internet, we note that the raw csv file can be reached from the link <https://raw.githubusercontent.com/allisonhorst/palmerpenguins/refs/heads/main/inst/extdata/penguins.csv>. Using this as the argument to `read_csv()`, we do:

```{r}
read_csv("https://raw.githubusercontent.com/allisonhorst/palmerpenguins/refs/heads/main/inst/extdata/penguins.csv")
```

Alternatively, a copy of the csv file is included in this week's github repo under the `data` folder. After that, we can load the file via:

```{r}
read_csv("data/penguins.csv")
```

As with other data structure, you can assign a dataframe to a variable for future use and manipulation. In our case, we may do:

```{r}
penguins <- read_csv("data/penguins.csv")
```

### Inspecting a dataframe

Unlike csv or other pure text data format, each column of a dataframe has a specific data type. To inspect that, use the `glimpse()` function from `dplyr`, which also display useful information such as the number of rows and columns, as well as a preview of the data

```{r}
glimpse(penguins)
```

In addition, if you just want to sample some rows from the dataframe, you can use `head()`, `tail()`, and (with the `car` package) `some()`

```{r}
head(penguins)
```

```{r}
tail(penguins)
```

```{r}
some(penguins)
```

To get the number of rows and columns programmatically (i.e., assignable to variables and thus can be further manipulated), use `nrow()` and `ncol()`

```{r}
nr <- nrow(penguins)
nc <- ncol(penguins)

print(c(nr, nc))
```

### Column operations

Load some data from the `nycflight23` package

```{r}
library(nycflights23)
glimpse(flights)
```

Sample some rows for the dataframe to get a sense of the data

```{r}
some(flights)
```

#### `select()`

Extract only a few columns from the above dataframe, as store the result in `flights_sub`

Note that dataframes are "immutable". This means that you should treat `flights` and `flights_sub` as completely independent of each other.

```{r}
flights_sub <- select(flights, 
  c("carrier", "flight", "origin", "dest", "air_time", "distance"))
some(flights_sub)
```

Alternative syntax (often more convenient). Notice that column names are being treated like variable in this form

```{r}
flights_sub <- select(flights, carrier:flight | origin:distance)
some(flights_sub)
```

#### `mutate()`

Define a new column using `mutate()`

```{r}
flights_new <- mutate(flights_sub, speed = distance / air_time)
some(flights_new)
```

If the named argument coincide with the name of an existing column, the existing column is being overwritten instead

```{r}
flights_new <- mutate(flights_new, flight = as.character(flight))
some(flights_new)
```

#### `rename()` and `relocate()`

Rename a column:

```{r}
flights_new <- rename(flights_new, destination=dest)
some(flights_new)
```

reorder columns using relocate:

```{r}
flights_new <- relocate(flights_new, speed, .before = air_time)
some(flights_new)
```

Perform the above transformations on our dataframe in one go, and using the pipe operator `|>`

```{r}
flights_chained <- flights |> select(carrier:flight | origin:distance) |> 
  mutate(speed = distance / air_time) |>
  mutate(flight = as.character(flight)) |>
  rename(destination=dest) |>
  relocate(speed, .before = air_time)

some(flights_chained)
```

### Row operations

#### `filter()`

Filter rows of a dataframe

```{r}
flights |> filter(dep_delay > 240)
```

Filter rows of a dataframe using compound conditions

```{r}
flights |> filter(dep_delay > 240 & arr_delay > 240)
```

#### `arrange()`

Arrange the rows of a dataframe by the value of a single column

```{r}
flights |> arrange(carrier)
```

Arrange the rows of a dataframe by a cascading set of columns

```{r}
flights |> arrange(carrier, flight)
```

Display the most delayed flight by sorting departure delay in _descending_ order

```{r}
flights |> arrange(desc(dep_delay)) |> head()
```

#### `distinct()`

Extract unique combination of carrier and flight number

```{r}
flights |> distinct(carrier, flight)
```

Extract the first flight for each combination of carrier and flight number

```{r}
flights |> arrange(time_hour, minute) |>
  distinct(carrier, flight, .keep_all=TRUE)
```

### Saving dataframe to .csv file or .rds file

Create the dataframe to be saved

```{r}
flights_LGA_delay <- flights |> 
  select(carrier:flight | origin:dest | dep_delay | arr_delay | time_hour) |>
  filter(origin == "LGA" & dep_delay > 240)
```

Export the dataframe as a .csv file

```{r}
write_csv(flights_LGA_delay, "flights_LGA_delay.csv")
```

Export the dataframe to R native .rds file

```{r}
write_rds(flights_LGA_delay, "flights_LGA_delay.rda")
```

Load the saved rds file

```{r}
filghts_from_rds <- read_rds("flights_LGA_delay.rda")
some(filghts_from_rds)
```
